Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2024-01-23 main.tex
:pencil: Update background

\begin{abstract}
	This paper investigates how a passive \gls{ris} can shape the \gls{mimo} point-to-point channel in terms of singular values.
	We depart from the widely-adapted diagonal phase shift model to a general \gls{bd} architecture, which provides superior shaping capability thanks to in-group connections between elements.
	An efficient \gls{rcg} algorithm is tailored for smooth optimization problems of asymmetric \gls{bd} \gls{ris} with arbitrary group size, then invoked for the Pareto frontier of channel singular values.
	To understand the gain from off-diagonal entries, we also derive analytical singular value bounds in \gls{los} and fully-connected scenarios.
	As a side product, we tackle \gls{mimo} rate maximization problem by alternating between active beamformer (eigenmode transmission) and passive beamformer (\gls{rcg} algorithm) until convergence.
	A low-complexity suboptimal solution that decouples both blocks is also proposed, where the channel shaping subproblem is formulated as power maximization and solved iteratively in closed form.
	Theoretical analysis and numerical evaluation reveal that the shaping advantage of \gls{bd} \gls{ris} increases with group size and stems from subchannel rearrangement and subspace alignment.
\end{abstract}

@@

\begin{section}{Introduction}
	% The quest for reliable, high-speed, and ubiquitous wireless connectivity has been long-standing since Marconi's illuminating radio in 1895.
	% Great successes have been made at transmitter and receiver sides over the past century, and the society is unprecedentedly close to the Shannon limit \cite{Shannon1948}.
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that serves on demand.
	This is empowered by the recent advances in \gls{ris}, a real-time programmable metasurface of numerous non-resonant sub-wavelength scattering elements.
	It can manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019} with a higher energy efficiency, lower cost, lighter footprint, and greater scalability than relays.
	Using \gls{ris} for {passive beamforming} has attracted significant interest in wireless communication \cite{Wu2019,Wu2020c,Yang2020,Zheng2021}, backscatter \cite{Jia2020,Liang2022}, sensing \cite{Liu2022a,Hua2023}, and power transfer literature \cite{Wu2021d,Feng2022,Zhao2022}, reporting a second-order array gain and fourth-order power scaling law (with proper waveform).
	On the other hand, \gls{ris} also enables {backscatter modulation} by dynamically switching between different patterns, as already investigated \cite{Karasik2020,Basar2020,Zhao2022a} and prototyped \cite{Tang2019a,Dai2020a}.
	Although fruitful outcomes have been harvested, one fundamental unanswered question is the {channel shaping} capability: \emph{To what extent can a passive \gls{ris} reshape the wireless channel?}

	The answer indeed depends on the hardware architecture and scattering model.
	In conventional (a.k.a. diagonal) \gls{ris}, each scattering element is tuned by a dedicated impedance and acts as an \emph{individual} phase shifter \cite{Wu2020}.
	% The concept is generalized to \gls{bd} \gls{ris} \cite{Shen2020a,Li2023b} featuring in-group connections between elements.
	The concept is generalized to \gls{bd} \gls{ris} \cite{Shen2020a,Li2023b} that connects adjacent elements as a group using passive components.
	This allows \emph{cooperative} scattering --- wave impinging on one element can propagate within the circuit and depart partially from any elements in the same group.
	\gls{bd} \gls{ris} can thus control both amplitude and phase of the reflected wave, generalizing the scattering matrix from diagonal with unit-magnitude entries to block diagonal with  unitary blocks.
	Its benefit has been recently shown in receive power maximization \cite{Nerini2023,Santamaria2023,Fang2023,Nerini2023a}, transmit power minimization \cite{Zhou2023}, and rate maximization \cite{Zhou2023,Nerini2023a,Li2023d,Bartoli2023,Li2023c}.
	The channel estimation \cite{Li2023e} and mutual coupling \cite{Li2023f} issues have also been investigated.
	Therefore, \gls{bd} \gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

	% Attempts to characterize the channel shaping capability can be classified into \emph{singular value centric} and \emph{power centric} methods.
	Channel shaping metrics can be classified into \emph{singular value centric} and \emph{power centric}.
	For diagonal \gls{ris}, the former has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023} in \gls{pc}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023} in \gls{ic}.
	The latter has been studied in terms of channel power \cite{Wu2019} in \gls{pc} and leakage interference \cite{Santamaria2023a} in \gls{ic}.
	For \gls{bd} \gls{ris}, the only metric analyzed is the channel power \cite{Nerini2023,Santamaria2023} while the \gls{mimo} case is still unexplored.\footnote{Single-stream \gls{mimo} with given precoder and combiner was considered in \cite{Nerini2023}. In terms of channel shaping, it is equivalent to \gls{siso}.}
	Although insightful, those attempts are far from complete and the techniques are limited to specific scenarios.
	This paper aims for a comprehensive answer to the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.

>>>>

\begin{abstract}
	This paper investigates how a passive \gls{ris} can reshape the \gls{mimo} point-to-point channel in terms of singular values.
	We depart from the widely-adapted diagonal phase shift model to a general \gls{bd} architecture, which provides superior shaping capability thanks to in-group connections between elements.
	An efficient \gls{rcg} algorithm is tailored for smooth optimization problems of asymmetric \gls{bd} \gls{ris} with arbitrary group size, then invoked for the Pareto frontier of channel singular values.
	To understand the gain from off-diagonal entries, we also derive analytical singular value bounds in \gls{los} and fully-connected scenarios.
	As a side product, we tackle \gls{mimo} rate maximization problem by alternating between active beamformer (eigenmode transmission) and passive beamformer (\gls{rcg} algorithm) until convergence.
	A low-complexity suboptimal solution decoupling both blocks is also proposed, where the channel shaping subproblem is formulated as channel power maximization and solved iteratively in closed form.
	Theoretical analysis and numerical evaluation reveal that the shaping advantage of \gls{bd} \gls{ris} increases with group size and stems from stronger subchannel rearrangement and subspace alignment capabilities.
\end{abstract}

@@

\begin{section}{Introduction}
	% The quest for reliable, high-speed, and ubiquitous wireless connectivity has been long-standing since Marconi's illuminating radio in 1895.
	% Great successes have been made at transmitter and receiver sides over the past century, and the society is unprecedentedly close to the Shannon limit \cite{Shannon1948}.
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that serves on demand.
	This is empowered by the recent advances in \gls{ris}, a real-time programmable metasurface of numerous non-resonant sub-wavelength scattering elements.
	It can manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019} with a higher energy efficiency, lower cost, lighter footprint, and greater scalability than relays.
	Using \gls{ris} for {passive beamforming} has attracted significant interest in wireless communication \cite{Wu2019,Wu2020c,Yang2020,Zheng2021}, backscatter \cite{Jia2020,Liang2022}, sensing \cite{Liu2022a,Hua2023}, and power transfer literature \cite{Wu2021d,Feng2022,Zhao2022}, reporting a second-order array gain and fourth-order power scaling law (with proper waveform).
	On the other hand, \gls{ris} also enables {backscatter modulation} by dynamically switching between different patterns, as already investigated \cite{Karasik2020,Basar2020,Zhao2022a} and prototyped \cite{Tang2019a,Dai2020a}.
	Despite fruitful outcomes, one critical unanswered question is the {channel shaping} capability: \emph{To what extent can a passive \gls{ris} reshape the wireless channel?}

	The answer indeed depends on the hardware architecture and scattering model.
	In conventional (a.k.a. diagonal) \gls{ris}, each scattering element is tuned by a dedicated impedance and acts as an \emph{individual} phase shifter \cite{Wu2020}.
	The concept is generalized to \gls{bd} \gls{ris} \cite{Shen2020a,Li2023b} which groups adjacent elements using passive components.
	This allows \emph{cooperative} scattering --- wave impinging on one element can propagate within the circuit and depart partially from any element in the same group.
	\gls{bd} \gls{ris} can thus control both amplitude and phase of the reflected wave, generalizing the scattering matrix from diagonal with unit-magnitude entries to block diagonal with  unitary blocks.
	Its benefit has been recently shown in receive power maximization \cite{Nerini2023,Santamaria2023,Fang2023,Nerini2023a}, transmit power minimization \cite{Zhou2023}, and rate maximization \cite{Zhou2023,Nerini2023a,Li2023d,Bartoli2023,Li2023c}.
	Practical issues such as channel estimation \cite{Li2023e} and mutual coupling \cite{Li2023f} have also been investigated.
	Therefore, \gls{bd} \gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

	% Attempts to characterize the channel shaping capability can be classified into \emph{singular value centric} and \emph{power centric} methods.
	% Channel shaping is different from passive beamforming as it seeks to modify the inherent properties of the channel itself, allowing one to decouple \gls{ris} and transceiver design.
	Channel shaping is different from passive beamforming as it seeks to modify the inherent properties of the channel itself.
	This allows one to decouple the \gls{ris}-transceiver design and explore the fundamental limits of channel manipulation.
	For example, diagonal \gls{ris} has been proved useful for improving channel power \cite{Ning2020}, degree of freedom \cite{Ozdogan2020,Li2023h}, condition number \cite{Zheng2022,Huang2023}, and effective rank \cite{ElMossallamy2021,Meng2023} in \gls{mimo}.
	On the other hand, \gls{bd} \gls{ris} can provide a higher channel power but the results are limited to \gls{siso} \cite{Nerini2023} and \gls{miso} \cite{Santamaria2023}.
	While these studies offer promising glimpses into the channel shaping potential of passive \gls{ris}, a comprehensive understanding of its capabilities and limitations is desired, and a universal design framework is still missing.
	% \footnote{In terms of channel shaping, single-stream \gls{mimo} with given precoder and combiner \cite{Nerini2023} is equivalent to \gls{siso}.}.
	% Although insightful, those attempts are far from complete and the techniques are limited to specific scenarios.
	This paper aims to answer the channel shaping question through theoretical analysis and numerical optimization.
	The contributions are summarized below.
